A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-21 FMCAD2021/4_dec_proc.tex
space adjustment

\subsection{Normalization}

@@

The normalization step first removes the occurrences of $\lambda_{10}(\iterm)$ and $\ell_{10}(\iterm)$  such that $\ivarx$ occurs  in $\iterm$, by encoding them with the exponential function. Moreover, for each occurrence of $10^\iterm$ such that $\ivarx$ occurs in $\iterm$, it introduces a fresh variable $\ivarz$, replaces $10^\iterm$ with $10^\ivarz$, and adds the equality $\ivarz = \iterm$.  It also applies some additional transformations. After the normalization, the resulting formula is of the following shape: 1) it is in NNF (negation normal form),  2) it contains no occurrences of $\lambda_{10}(\iterm)$ as well as $\ell_{10}(\iterm)$ such that $\ivarx$ occurs in $\iterm$, 3)  it contains no occurrences of $10^\iterm$ such that $\ivarx$ occurs in $\iterm$, but $\iterm \neq \ivarx$, 4) moreover, all the atomic formulas are of the form $\iterm_1 \le \iterm_2$ or $c | \iterm$. More details can be found in Appendix~\ref{app-norm}.

@@

\subsection{Elimination of  exponential occurrences of variables}\label{sec-elim-exp}

@@

Recall that after normalization, the atomic formulas are of the form $\iterm_1 \le \iterm_2$ or $c | \iterm$. Therefore, we can assume that the atomic formulas in $\varphi''_{\sigma,i}$ are  of the form 
%
$a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} \le \iterm(\vec{\ivary})$
or  
$c \ \big  | \ \big(a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} + \iterm(\vec{\ivary})\big).$
%
For convenience, let us call these formulas as inequality respectively divisibility atomic formulas. In the sequel, we illustrate how to eliminate the exponential occurrences of $\ivarx_{\sigma(i)}$ for these the inequality atomic formulas. The elimination of the exponential occurrences of the divisibility formulas are simpler and omitted, which can be found in Appendix~\ref{app-div}.

>>>>

\vspace*{-3mm}
\subsection{Normalization}

@@

The normalization step first removes the occurrences of $\lambda_{10}(\iterm)$ and $\ell_{10}(\iterm)$  such that $\ivarx$ occurs  in $\iterm$, by encoding them with the exponential function. Moreover, for each occurrence of $10^\iterm$ such that $\ivarx$ occurs in $\iterm$, it introduces a fresh variable $\ivarz$, replaces $10^\iterm$ with $10^\ivarz$, and adds the equality $\ivarz = \iterm$.  It also applies some additional transformations. After the normalization, the resulting formula is of the following shape: 1) it is in NNF (negation normal form),  2) it contains no occurrences of $\lambda_{10}(\iterm)$ as well as $\ell_{10}(\iterm)$ such that $\ivarx$ occurs in $\iterm$, 3)  it contains no occurrences of $10^\iterm$ such that $\ivarx$ occurs in $\iterm$, but $\iterm \neq \ivarx$, 4) all the atomic formulas are of the form $\iterm_1 \le \iterm_2$ or $c | \iterm$. More details can be found in Appendix~\ref{app-norm}.

@@

\vspace*{-3mm}
\subsection{Elimination of  exponential occurrences of variables}\label{sec-elim-exp}

@@

Recall that after normalization, the atomic formulas are of the form $\iterm_1 \le \iterm_2$ or $c | \iterm$. Therefore, we can assume that the atomic formulas in $\varphi''_{\sigma,i}$ are  of the form 
%
$a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} \le \iterm(\vec{\ivary})$
or  
$c \ \big  | \ \big(a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} + \iterm(\vec{\ivary})\big).$
%
For convenience, let us call these formulas as inequality respectively divisibility atomic formulas. In the sequel, we illustrate how to eliminate the exponential occurrences of $\ivarx_{\sigma(i)}$ for these inequality atomic formulas. The elimination of the exponential occurrences of the divisibility formulas are simpler and omitted, which can be found in Appendix~\ref{app-div}.

### 2021-05-21 FMCAD2021/5_optimization.tex
space adjustment

In this section, we focus on quantifier-free {\paexp} formulas (or existential {$\paexp$} formulas since they are satisfiability-equivalent), and present various optimizations of the decision procedure for {\paexp}, aiming at an efficient implementation. The focus on quantifier-free {\paexp} formulas is motivated by the following two facts: 1) the flattening of {$\strint$} string constraints results into such formulas, 2) these formulas are already challenging for the state-of-the-art SMT solvers e.g. CVC4 and Z3 (with exponential functions defined as recursive functions).

@@

Recall that in the decision procedure for {\paexp}, after the normalization, the orders of variables are enumerated and for each order, the exponential as well as linear occurrences of variables are eliminated. Since the quantifier elimination is expensive and applied to each possible order of variables, if we could reduce the candidate variable orders in the very beginning, it would facilitate considerable speed-up for the decision procedure.

@@

Recall that Lemma~\ref{lem:exp-ineq} enables us to eliminate the exponential occurrences of $\ivarx_{\sigma(i)}$ from 
$$
\begin{array}{l}
\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def  \\
\hspace{4mm} 
a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} \le c.
\end{array}
$$
Actually, Lemma~\ref{lem:exp-ineq} does more in the sense that all occurrences of $\ivarx_{\sigma(i)}$, including the linear ones, are eliminated from the atomic formulas resulted from $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, e.g. $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-1)} + p /\ivarx_{\sigma(i)}]$. Then we can continue eliminating the exponential occurrences of $\ivarx_{\sigma(i-1)}$ from $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-1)} + p /\ivarx_{\sigma(i)}]$, provided that the coefficient of $\ivarx_{\sigma(i-1)}$ therein is nonzero. Iterating this process would produce a {\pa} formula eventually.

@@

To avoid applying the quantifier elimination algorithm of {\pa}, we consider the following under approximation of $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, namely, we additionally assume that $\ivarx_{\sigma(i)} \le 10^{u}$ for some constant bound $u \in \Nat$ with $u \ge 1$. Then $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$ can be rewritten as 
%\begin{equation}\label{eqn-org}
\[\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def \sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c - b_i  \ivarx_{\sigma(i)}.\]
%\end{equation}
Let us assume $a_{i-1} > 0$.
Define $c_1, c_2$ as follows: If $b_i > 0$, then $c_1 = c- b_i 10^u$ and $c_2 = c$, otherwise, $c_1 = c$ and $c_2 = c - b_i 10^u$.
It is easy to observe that $c_1 \le c - b_i \ivarx_{\sigma(i)} \le c_2$.
Then we can apply Lemma~\ref{lem:exp-ineq} to the following two inequalities to eliminate $10^{\ivarx_{\sigma(i-1)}}$,
\begin{equation}\label{eqn-lb}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_1
\end{equation}
and
\begin{equation}\label{eqn-ub}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_2.
\end{equation}
Let $\alpha_1 = \ell_{10}(c_1) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c_2) -  \ell_{10}(a_{i-1})$. Then from Lemma~\ref{lem:exp-ineq}, 
\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-lb}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\ltrue$,
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-ub}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\lfalse$.
\end{itemize}
Therefore, $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, thus also $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is equivalent to 
\[
\small
\begin{array}{l}
\bigvee \limits_{p=0}^{B-1} ( \ivarx_{\sigma(i-1)} = p \wedge \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) [p/\ivarx_{\sigma(i-1)}])   \\
\bigvee \big(\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  - 1  \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} +\delta \big)\\
%
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  -1 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \ge \alpha_2 + 2 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p = \alpha_1}^{\alpha_2 + 1}
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} = p\ \wedge \\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right),
\end{array}
\]
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated.

>>>>

In this section, we focus on quantifier-free {\paexp} formulas (or existential {\paexp} formulas since they are satisfiability-equivalent), and present various optimizations of the quantifier elimination procedure for {\paexp}, aiming at an efficient implementation. The focus on quantifier-free {\paexp} formulas is motivated by the following two facts: 1) the flattening of {$\strint$} constraints results into such formulas, 2) these formulas are already challenging for state-of-the-art SMT solvers (with exponential functions defined as recursive functions).

@@

Recall that in the decision procedure for {\paexp}, after the normalization, the variable orders are enumerated and for each order, the exponential and linear occurrences of variables are eliminated. Since the quantifier elimination is expensive and applied to each possible order of variables, if we could reduce the candidate variable orders in the very beginning, it would facilitate considerable speed-up for the decision procedure.

@@

Recall that Lemma~\ref{lem:exp-ineq} enables us to eliminate the exponential occurrences of $\ivarx_{\sigma(i)}$ from 
\vspace{-1mm}
$$
\begin{array}{l}
\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def  \\
\hspace{4mm} 
\vspace{-1mm}
a_i 10^{\ivarx_{\sigma(i)}}+\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i}b_k \ivarx_{\sigma(k)} \le c.
\end{array}
$$
Actually, Lemma~\ref{lem:exp-ineq} does more in the sense that all occurrences of $\ivarx_{\sigma(i)}$, including the linear ones, are eliminated from the atomic formulas resulted from $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, e.g. $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-1)} + p /\ivarx_{\sigma(i)}]$. Then we can continue eliminating the exponential occurrences of $\ivarx_{\sigma(i-1)}$ from $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-1)} + p /\ivarx_{\sigma(i)}]$, provided that the coefficient of $\ivarx_{\sigma(i-1)}$ therein is nonzero. Iterating this process would produce a {\pa} formula eventually.

@@

To avoid applying the quantifier elimination algorithm of {\pa}, we consider the following under approximation of $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, namely, we additionally assume that $\ivarx_{\sigma(i)} \le 10^{u}$ for some constant bound $u \in \Nat$ with $u \ge 1$. Then $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$ can be rewritten as 
%\begin{equation}\label{eqn-org}
\vspace{-2mm}
\[\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) \Def \sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c - b_i  \ivarx_{\sigma(i)}.\]
\vspace{-1mm}
%\end{equation}
Let us assume $a_{i-1} > 0$.
Define $c_1, c_2$ as follows: If $b_i > 0$, then $c_1 = c- b_i 10^u$ and $c_2 = c$, otherwise, $c_1 = c$ and $c_2 = c - b_i 10^u$.
It is easy to observe that $c_1 \le c - b_i \ivarx_{\sigma(i)} \le c_2$.
Then we can apply Lemma~\ref{lem:exp-ineq} to the following two inequalities to eliminate $10^{\ivarx_{\sigma(i-1)}}$,
\vspace{-2mm}
\begin{equation}\label{eqn-lb}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_1
\end{equation}
and
\vspace{-2mm}
\begin{equation}\label{eqn-ub}
\sum_{j=1}^{i-1} a_j 10^{\ivarx_{\sigma(j)}} + \sum_{k=1}^{i-1}b_k \ivarx_{\sigma(k)} \le c_2.
\end{equation}
Let $\alpha_1 = \ell_{10}(c_1) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c_2) -  \ell_{10}(a_{i-1})$. Then from Lemma~\ref{lem:exp-ineq}, 
\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-lb}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\ltrue$,
\item if $\ivarx_{\sigma(i-1)} \ge B$, $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-ub}), thus also $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is evaluated to $\lfalse$.
\end{itemize}
Therefore, $\tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, thus also $\tau(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})$, is equivalent to 
\[
\small
\begin{array}{l}
\bigvee \limits_{p=0}^{B-1} ( \ivarx_{\sigma(i-1)} = p \wedge \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)}) [p/\ivarx_{\sigma(i-1)}])   \\
\bigvee \big(\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  - 1  \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} +\delta \big)\\
%
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \le \alpha_1  -1 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p=0}^{\delta-1} 
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} \ge \alpha_2 + 2 \ \wedge \\
 \ivarx_{\sigma(i-1)} = \ivarx_{\sigma(i-2)} +p\ \wedge\\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[\ivarx_{\sigma(i-2)} + p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right)\\
\bigvee \bigvee \limits_{p = \alpha_1}^{\alpha_2 + 1}
\left(
\begin{array}{l}
\ivarx_{\sigma(i-1)} \ge B \wedge \ivarx_{\sigma(i-1)} = p\ \wedge \\
 \tau'(\ivarx_{\sigma(i)}, \ldots, \ivarx_{\sigma(1)})[p /\ivarx_{\sigma(i-1)}] 
\end{array}
\right),
\end{array}
\]
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated.
